---
title: "C'est une pipe."
description: |
  What exactly is Magrittr's Pipe? 
author:
  - name: Logan Piepmeier
    url: {}
date: 2021-11-17
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
```

## What is a pipe in R?

The `%>%` pipe has become pretty ubiquitous in R, especially with the rise of the [Tidyverse](https://www.tidyverse.org/) packages. If you're not familiar with it, just know that a pipe allows you to write your commands in a different (and often more natural) order. Using dplyr's `filter` function on the well-known `iris` data set: `filter(iris, species == "Setosa")` can be rewritten `iris %>% filter(species == "Setosa")` using a pipe. This may not seem more natural at first, but its utility becomes more obvious when writing chains of several commands.

For another dplyr / `iris` example, we may want to use dplyr to find the mean `Sepal.Width` to `Sepal.Length` ratio for each species in the `iris` data set. We could do this in a single command with:

```{r, echo = TRUE, message=FALSE}
summarize(group_by(mutate(iris, sepal_ratio = Sepal.Width / Sepal.Length), Species), mean_sepal_ratio = mean(sepal_ratio))
```

Even if you know dplyr well, it likely took a few passes to understand that line. Pipes (IMHO) make this much more readable.

```{r, echo = TRUE}
iris %>%
  mutate(sepal_ratio = Sepal.Width / Sepal.Length) %>%
  group_by(Species) %>%
  summarize(mean_sepal_ratio = mean(sepal_ratio))
```

Pipes let you write functions in the order you might verbally explain what the code is doing. That's useful!

This pattern in R is borrowed from other, more object-oriented, programming languages' concept of "method chaining". R's pipes aren't quite the same thing, because we're still in the world of functional rather than object-oriented programming, but maybe I'll return to that difference in another post.

## No, but what *is* a pipe in R?

`%>%`, like the familiar `%in%`, is a *special* function (and by that I mean that's what it's called). In particular, both `%in%` and `%>%` are "infix functions". Like their linguistic namesake, they go in between objects to modify the meaning of a statement. In English, infix is almost always impolite or informal. Think "Absolutely" becoming "abso-fucking-lutely!" and "another" becoming "a-whole-nother".

Staying with that familiar `%in%` for a moment, you can actually see that it's secretly a normal old function under the hood. Here is the typical use of `%in%`:

```{r, echo = TRUE}
c(1,2,3) %in% c(2,3,4)
```

But we can also use it this way:

```{r echo = TRUE}
`%in%`(c(1,2,3), c(2,3,4))
```

So `%in%` can actually be used with the `function(x,y)` syntax we're used to. Play around with R and you'll find many very basic operations work this way when wrapped in backticks:

```{r, echo = TRUE}
`+`(1,1) # Addition
```

```{r, echo = TRUE}
`%%`(5, 3) # Modulo
```

```{r, echo = TRUE}
`^`(2, 3) # Exponentiation 
```

You can even define your own infix function by defining a two-argument function and wrapping its name in "% %". Just use the backticks when defining it.

```{r, echo = TRUE}
`%my_very_own_plus_sign%` <- function(x,y){ x + y }
2 %my_very_own_plus_sign% 2
```

*As an aside, R lets you define infix functions with any number of arguments. They just ignore any additional arguments if you provide more than two, or crash on execution if you define fewer.*

As you've surely guessed by now, you can also call the pipe with this syntax (although its pretty clunky to do so).

```{r, echo = TRUE}
`%>%`('this is', paste('my example'))
```

There's no reason to actually call infix functions this way. That defeats their purpose as convenient, more readable ways to structure your R code.

## Yeah but what *is* a pipe *really*?

Ok, so the `%>%` pipe is an infix function. But I still don't know how it actually does what it does! And really, as I write this, I have no idea how it works. We are in this together.

Well, at a basic level, we can just ask R what it is.

```{r, echo = TRUE}
`%>%`
```

Let's go line-by-line.

-   It's a function of two variables -- we knew that.

-   The function calls `substitute` on each argument. According to the docs, `substitute` "returns the parse tree for the (unevaluated) expression `expr`, substituting any variables bound in `env`". That's... not helpful. I've explored a bit and found `substitute` returns a name object for its argument.

-   It sets `kind` to integer 1. But... `kind` doesn't seem to be referenced again.

-   `env` is `parent.frame`. I remember `parent.frame` from [Hadley's Advanced R](https://adv-r.hadley.nz/environments.html?q=parent.frame#call-stack). It refers to the environment of variable outside of the function (in the environment where the function was called). However, this also doesn't get referred to again.

-   `lazy <- TRUE`. I mean, I get that. But again, `lazy` doesn't show up again.

-   `.External2(magrittr_pipe)`. I think we have reached the crux of this problem. There does not appear to be a variable called `magrittr_pipe` in this function, and `??magrittr_pipe` doesn't have any results in R. Calling `help(.External2)` points to `.External` (thanks, R docs). `help(.External)` tells me that this function is "to pass R objects to compiled C/C++ code that has been loaded into R."

Uh oh, C code. To a high-level coder like me, this is the part of the map that reads "here be dragons". I hope to be brave enough to explore this territory in a future post.
